New insights into the Weddell Sea ecosystem applying a quantitative network approach

### 2023-11-09 Manuscript/Manuscript.Rmd
Revised version to submit

---
title: New insights into the Weddell Sea ecosystem applying a quantitative network approach
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'communication')`"
author:
  - given_name: Tomás I.
    surname: Marina
    affiliation: "1, *"
    email: tomasimarina@gmail.com
  - given_name: Leonardo A.
    surname: Saravia
    affiliation: "1, *"
    email: arysar@gmail.com
  - given_name: Susanne
    surname: Kortsch
    affiliation: 2
correspongdingauthors: Tomás I. Marina (tomasimarina@gmail.com) and Leonardo A. Saravia (arysar@gmail.com)
affiliation:
  - code: 1
    address: Centro Austral de Investigaciones Científicas (CADIC-CONICET), Ushuaia, Argentina
  - code: 2
    address: Spatial Foodweb Ecology Group, Department of Agricultural Sciences, University of Helsinki, Finland
  - code: "*"
    address: These authors contributed equally to this work.
abstract: |
  Network approaches can shed light on the structure and stability of complex marine communities. In recent years, such approaches have been successfully applied to study polar ecosystems, improving our knowledge on how they might respond to ongoing environmental changes. The Weddell Sea is one of the most studied marine ecosystems outside the Antarctic Peninsula in the Southern Ocean. Yet, few studies consider the known complexity of the Weddell Sea food web, which in its current form comprises 490 species and 16041 predator-prey interactions. Here we analysed the Weddell Sea food web, focusing on the species and trophic interactions that underpin ecosystem structure and stability. We estimated the strength for each interaction in the food web, characterised species position in the food web using unweighted and weighted food web properties, and analysed species’ roles with respect to the stability of the food web. We found that the distribution of the interaction strength (IS) at the food web level is asymmetric, with many weak interactions and few strong ones. We detected a positive relationship between species mean IS and two unweighted properties (i.e., trophic level and the total number of interactions). We also found that only a few species possess key positions in terms of food web stability. These species are characterised by high mean IS, mid to high trophic level, relatively high number of interactions, and mid to low trophic similarity. In this study, we integrated unweighted and weighted food web information, enabling a more complete assessment of the ecosystem structure and function of the Weddell Sea food web. Our results provide new insights, which are important for the development of effective policies and management strategies, particularly given the ongoing initiative to implement a Marine Protected Area (MPA) in the Weddell Sea.

@@

Estimating interaction strength (IS) in food webs allows differentiating the importance of species interactions. On the contrary, unweighted food web representations give equal importance to all interactions well-knowing that some species interactions are stronger than others and hence play a different role for ecosystem functioning and stability. Both empirical and theoretical studies show that interactions strength distributions in food webs are assymmetric [@Paine1992; @McCann1998; @Emmerson2004; @Wootton2005; @Kortsch2021], containing a few strong and many weak interactions. This asymmetric patterning of weak and strong links is crucial to food web stability [@Paine1992; @McCann1998; @Neutel2002]. In a recent paper on an aquatic food web it was further highlighted that temporal changes in ecosystem functioning could only be predicted using weighted food web structure [@Kortsch2021]. Hence, in order to assess the stability and functioning of a food web, it is important to first determine the quantitative structure of the trophic network.

Several methodologies have been applied to estimate IS in food webs, where the quantity and quality of the data mostly determines which approach is the most convenient [@Berlow2004]. Approaches include experimental methods combined with dynamic modelling [@Emmerson2004; @Carrara2015], measurements of species abundances through time [@Fahimipour2014; @Chang2021], and estimation of metabolic rates and biomass of all species in the community [@Neutel2014]. However, these types of methods require large experimental set-ups and parameterisations restricting the analyses to smaller networks (e.g., approximately 10 species or less). Other methods based on allometric scaling relationships and biomass information [@Kortsch2021; @Gauzens2019] can be applied to larger networks with less data requirements, but this comes at the expense of precision in the predictions. For even larger food webs composed of nearly 1000 species and more than 10000 interactions, only methods with even less data requirements are feasible. One of these methods, proposed by @Pawar2012, combines data on consumer and resource body masses and consumer search space (interaction dimensionality) to obtain IS estimates for each pairwise predator-prey interaction. An advantage of this method is that it can be applied without information on species biomass.

Using a a network approach, different types (e.g., terrestrial, lake, marine) of food webs from various geographic locations are studied worldwide, including marine polar food webs [@Carscallen2012; @Santana2013; @Kortsch2019; @Pecuchet2022]. Some of the studies from the Arctic show how food web properties (e.g., connectance) are constrained by environmental factors such as sea ice cover and seawater temperature [@Kortsch2019; @Pecuchet2022]. In the Southern Ocean, important insights have been gained into mechanisms of energy flow, the relative importance of individual species and their traits, and the influence of environmental variables (e.g., sea-ice) on the structure of local food webs [@Cordone2020; @Rossi2019]. For instance, in Potter Cove (West Antarctic Peninsula) the substratum type (i.e. hard/soft or rocks/sediments) plays a significant role in the structure and stability of the food web. In Terra Nova Bay (Ross Sea), the architecture of biodiversity was reshaped by the pulsed input of sympagic food sources following sea-ice break up, with food web simplification, decreased intraguild predation, potential disturbance propagation and increased vulnerability to biodiversity loss [@Rossi2019].

The Weddell Sea is expected to be one of the last regions of the Southern Ocean to experience the consequences of climate change due to its extensive ice cover and ocean currents [@Teschke2021] resulting in less sea surface warming compared to other areas of the Southern Ocean. This Sea plays an important role in driving global thermohaline circulation and ventilating the global abyssal ocean because it generates a considerable part of the Antarctic Bottom Water [@Fahrbach2009]. Because of these environmental characteristics, the Weddell Sea may serve as a refuge for Antarctic species which depend on sea ice (e.g. krill, emperor penguin, Weddell seal) or have low heat tolerance (e.g., most notothenoid fishes) due to their adaptations to freezing temperatures [@Griffiths2017]. While essential large-scale hydrodynamic relationships are relatively well-known for this region [@deSteur2019], information on the current distribution, abundance and sensitivity to climate change is only partially known for a few species (e.g., emperor penguin) [@Houstin2022].

The network complexity of the Weddell Sea food web is high comprising 488 species and 16200 predator-prey interactions [@Jacob2011]. In an attempt to better understand species roles related to food web stability, Jacob et al. (2011) performed secondary extinction experiments and found that the removal of small to medium-sized, and not large, organisms caused a cascade of secondary extinctions. This findings highlighted the relative importance of predators, rather than prey, for the architecture, functioning and stability of the Weddell Sea food web, which coincides with findings from recent meta-analyses in natural complex food webs [@Brose2019; @Perkins2022]. Other investigations considered this food web in a meta-analysis context showing that high predator-prey body-mass ratios are found for predator groups with specific trait combinations, including small vertebrates and large swimming or flying predators [@Brose2019]. These trait combinations generate weak interactions that stabilize communities against perturbations maintaining ecosystem functioning.

In this study, we aim to go beyond a purely topological (presence/absence) assessment of who eats whom in the Weddell Sea ecosystem by providing a quantitative analysis of the trophic interaction network. We aim to analyse the species' role for the structure and stability of the food web. To achieve this, we: 1) estimated the strength for each interaction in the Weddell Sea food web, 2) characterised species' role considering both weighted and unweighted properties, and 3) analysed the species’ role related to the stability of the food web. This is the first time that interaction strengths were estimated for all pairwise trophic interactions at the species level (except for a few) for the Weddell sea food web.

@@

The high Antarctic Weddell Sea shelf is situated between 74 and 78ºS, stretching approximately 450 km from East to West (Figure 1). Water depth varies between 200 and 500 meters, and shallower areas are covered by continental ice, which forms the coastline along the eastern and southern part of the Weddell Sea. The shelf area contains a complex three-dimensional benthic habitat with large benthic biomasses, intermediate to high diversity in comparison to benthic boreal communities and a spatially patchy distribution of organisms [@Dayton1990; @Teixido2002].

@@

The Weddell Sea food web was retrieved from the GlobAL daTabasE of traits and food Web Architecture (GATEWAy, version 1.0) of the German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig [@Brose2018]. In addition to predator-prey interactions, the database contains information on other biological data such as the mean body mass and movement type for each species in the food web. Furthermore, it incorporates information about the interaction itself, such as the dimension of the predator search space (2 or 3 dimensions). In its current form, the Weddell Sea food web comprises 490 species and 16041 predator-prey interactions and constitutes one of the most resolved food webs constructed to date [@Jacob2011].

@@

To estimate the strength of each pairwise interaction in the food web, we followed an approach proposed by @Pawar2012. The minimum data requirements are body mass of the consumer (predator) and resource (prey), and the interaction dimensionality (ID) classified as 2 or 3 dimensions. The ID is defined as the search space dimensions of the predator, which is also equivalent to  the movement space of the prey. Thus, the ID is classified as 2D when both predator and prey move in 2D (e.g., both are benthic) or if a predator moves in 3D and a prey in 2D (e.g., pelagic predator on benthic prey). The ID is classified as 3D when both predator and prey move in 3D (e.g., both pelagic) or if the predator moves in 2D and the prey in 3D (e.g., benthic predator, pelagic prey) [@Pawar2012]. GATEWAy v.1.0 provides information on the mean body mass for consumers and resources, except for ‘detritus’ and ‘sediment’, and the dimensionality for the majority of the interactions, though the latter is missing in some cases (924 interactions). To complete the missing data on species ‘dimensionality’, we used information about the movement type of predators and prey included in GATEWAy.

@@

We obtained estimates for resource density and the search rate from the scaling relationships with the resource and the consumer mass, respectively [@Pawar2012]. The coefficients of such relationships, determined by ordinary least squares regression, vary with the interaction dimensionality. On one hand, resource density scales with resource mass as power-law with different exponents in 2D and in 3D. Since mean mass for resources ‘phytodetritus’ and ‘sediment’ were not available in GATEWAy, we considered the body mass of the smallest phytoplankton species (‘Fragilariopsis cylindrus’) as a proxy. This is justified by the fact that ‘phytodetritus’ and ‘sediment’ are mainly composed of dead or senescent phytoplankton reaching the seabed [@Wolanski2011]. On the other hand, search rate scales with consumer mass as power-law with exponents in 2D and in 3D.

We fitted six candidate models (Exponential, Gamma, log-Normal, Normal, Power-law and Uniform) the IS distribution using maximum likelihood [@McCallum2008], and selected the best fitting model by computing the Akaike Information Criterion AIC [@Burnham2002].

@@

To characterise the role of each species in the food web, we considered unweighted and weighted food web properties (Figure 2). Unweighted properties are related to properties commonly used in qualitative food web studies and only describe the presence or absence of interactions without any information on strength between a pairwise species link [@Martinez1991; @Dunne2002; @Borrelli2014]. In contrast, weighted properties capture the importance of a trophic interaction by considering its strength.

To assess species roles as a function of the weighted food web, we focused on mean IS defined as the average strength of all interactions for a given species. Further we calculated three unweighted species properties: a) species degree, i.e., the sum of in- and out-going interactions ; b) trophic level ; and c) trophic similarity, i.e., the trophic overlap based on shared and unique resources and consumers. These metrics were chosen to assess a species role based on the unweighted network. The species degree has often been equated with species importance to the structure and functioning within a food web, i.e. perturbations to high-degree species may therefore have more significant effects on the food web robustness to perturbations than low-degree species [@Dunne2002a; references in @Cirtwill2018a]. The trophic level offers information about how important a species is to its biotic community, i.e., top predators and primary producers are expected to have particularly large effects on the rest of their communities through top-down and bottom-up control, respectively [references in @Cirtwill2018a]. Trophic similarity is an index of trophic overlap considering the set of prey and predators for a pair of species; it measures one of the most important aspects of species’ niches, the trophic niche, and functional aspects of biodiversity [@Martinez1991; @Williams2000].

Furthermore, we took species habitat affiliation into account, which describes the physical position of a species within the ecosystem. Species were categorised as: 1) benthic, if a species lives on the seafloor; 2) pelagic, if a species lives close to the surface; 3) benthopelagic, if it moves between and connects the aforementioned environments; 4) demersal, if it lives and feeds on, or near, the bottom of the sea; and 5) land-based, if the consumer is not strictly aquatic but feeds predominantly on marine species. Species habitat affiliations were retrieved from @Jacob2011.

To study the relationship between species mean IS (weighted property) and the unweighted species properties, we performed linear regression analyses between the log mean IS and each of the aforementioned unweighted properties. Thus, we considered the IS as the dependent variable and the given unweighted property as the independent variable, and obtained the coefficients (slope and intercept) for the linear model. Models were fitted using the least squares approach. We also explored the mean IS distribution with the species habitat.

@@

To analyse the impact of species on food web stability, we performed extinction simulations deleting one species at a time, that is for every extinction, network size was reduced by one species only. After each extinction, we calculated the stability of the network minus the removed species (489 nodes) and compared it with that of the whole network (490 nodes in total). To calculate stability, we used the mean of the real part of the maximum eigenvalue of the Jacobian matrix using randomized Jacobians, keeping the predator-prey sign structure fixed [@Allesina2008; @Grilli2016]. This stability index indicates a more stable food web when it is negative. We performed 1000 simulations for each species removal and obtained a mean maximum eigenvalue for each case. Finally, we statistically analysed this difference with an Anderson-Darling test considering a p-value < 0.01 [@Scholz1987]. If the difference between the networks without and with the removed species is positive, then the stability of the whole food web is higher without the targeted species, in other words this species makes the network less stable. If the difference is negative, then the stability of the whole food web is lower without the targeted species, i.e., this species ahs a stabilizing effect. A detailed description on the stability calculations can be found in the supplementary material.

To identify the species with the highest effect on food web stability and their characteristics, we plotted the results of each species' extinction and its effect on food web stability (the stability difference) against weighted (interaction strength), and the three unweighted properties (trophic level, degree, and trophic similarity), and  species habitat affiliation.

@@

The statistical distribution that best fitted the empirical interaction strength distribution of the Weddell Sea food web was a ‘gamma’ due to the high proportion of weak interactions and the existence of a few strong interactions (Figure 3, Table S3).

@@

We found that the species’ mean IS (weighted property) shows different relationships with the unweighted properties analysed (Figure 4A-D). In this regard, there is a positive relationship between IS and trophic level, i.e., the higher the trophic level of the species, the higher its mean IS. We also found a significant but less evident positive relationship with species degree. Contrary, there was no significant relationship between mean IS and trophic similarity. Considering species habitat affiliation, the “Benthopelagic” and “Pelagic” categories contained the two species with the highest mean IS, the killer whale Orcinus orca and the colossal squid Mesonychoteuthis hamiltoni, respectively. However, the majority of the species with relatively higher IS belonged to the “Demersal” and “Land-based” habitats groups. Species inhabiting the benthic realm showed the lowest mean IS (Figure 4D).

@@

Our extinction analyses showed that the majority of species had no significant impact on food web stability after being removed (Figure 5). Most of the species (black points in figure 5) did not change the stability of the network considerably after being removed, except for a few species (red points in figure 5). Only 15 out of 490 species (3.06%) gave rise to significant changes in the food web’s stability after their removal (Table 2). Network stability increased after the removal of most of these species, i.e., these species have a negative effect on stability. Only two species significantly decreased network stability after being removed, the demersal fish Pagetopsis macropterus and the benthopelagic amphipod Maxilliphimedia longipes.

After exploring the stability difference against the species properties (Figure 5), we found that the species that generated a significant impact on the stability of the food web were characterised by: 1) high mean IS; 2) mid to high trophic levels (TL > 3.2); 3) relatively high number of interactions (Degree > 25); and 4) mid to low trophic similarity (TS < 0.16). Habitat wise, species with a significant impact on the stability were present in all habitats, except for the benthic realm. Table 2 shows the results for the species with highest impact on the food web stability.

@@

Our analyses show that the distribution of species IS at the network level is asymmetric, i.e., the Weddell Sea food web contains many weak interactions and only a few strong ones. This finding is consistent with many previous theoretical and empirical studies [e.g. @McCann1998; @Neutel2002; @Emmerson2004; @Wootton2005; @Kortsch2021]. The asymmetric distribution of IS in food webs has been interpreted as an explanation for the persistence of complex communities in nature [@Bascompte2005; @Allesina2015; @Nilsson2016]. Here we show that this pattern is also prevalent in the Weddell Sea, one of the most complex food webs to date, comprising 490 species and 16041 predator-prey interactions. This finding is in someway validating the method we used, validation of allometric methods of IS estimation not including interaction dimensionality has been performed for microcosmos of relatively few species [@Jonsson2018].

@@

```{r, out.width = "12cm", echo = FALSE, fig.cap = "Frequency distribution of (median) interaction strengths for the Weddell Sea food web. The distribution was best fitted to a ‘log-Normal’ model". Total number of interactions = 16041."}
knitr::include_graphics("Fig3_IntDist_sim.png")
```

>>>>

---
title: New insights into the Weddell Sea ecosystem applying a quantitative network approach
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'communication')`"
author:
  - given_name: Tomás I.
    surname: Marina
    affiliation: "1, *"
    email: tomasimarina@gmail.com
  - given_name: Leonardo A.
    surname: Saravia
    affiliation: "1, *"
    email: arysar@gmail.com
  - given_name: Susanne
    surname: Kortsch
    affiliation: 2
correspongdingauthors: Tomás I. Marina (tomasimarina@gmail.com) and Leonardo A. Saravia (arysar@gmail.com)
affiliation:
  - code: 1
    address: Centro Austral de Investigaciones Científicas (CADIC-CONICET), Ushuaia, Argentina
  - code: 2
    address: Spatial Foodweb Ecology Group, Department of Agricultural Sciences, University of Helsinki, Finland
  - code: "*"
    address: These authors contributed equally to this work.
abstract: |
  Network approaches can shed light on the structure and stability of complex marine communities. In recent years, such approaches have been successfully applied to study polar ecosystems, improving our knowledge on how they might respond to ongoing environmental changes. The Weddell Sea is one of the most studied marine ecosystems outside the Antarctic Peninsula in the Southern Ocean. Yet, few studies consider the known complexity of the Weddell Sea food web, which in its current form comprises 490 species and 16041 predator-prey interactions. Here we analysed the Weddell Sea food web, focusing on the species and trophic interactions that underpin ecosystem structure and stability. We estimated the strength for each interaction in the food web, characterised species position in the food web using unweighted and weighted food web properties, and analysed species’ roles with respect to the stability of the food web. We found that the distribution of the interaction strength (IS) at the food web level is asymmetric, with many weak interactions and few strong ones. We detected a positive relationship between species median IS and two unweighted properties (i.e., trophic level and the total number of interactions). We also found that only a few species possess key positions in terms of food web stability. These species are characterised by high median IS, mid to high trophic level, relatively high number of interactions, and mid to low trophic similarity. In this study, we integrated unweighted and weighted food web information, enabling a more complete assessment of the ecosystem structure and function of the Weddell Sea food web. Our results provide new insights, which are important for the development of effective policies and management strategies, particularly given the ongoing initiative to implement a Marine Protected Area (MPA) in the Weddell Sea.

@@

Estimating interaction strength (IS) in food webs allows differentiating the importance of species interactions. On the contrary, unweighted food web representations give equal importance to all interactions well-knowing that some species interactions are stronger than others and hence play a different role for ecosystem functioning and stability. Both empirical and theoretical studies show that interactions strength distributions in food webs are asymmetric [@Paine1992; @McCann1998; @Emmerson2004; @Wootton2005; @Kortsch2021], containing a few strong and many weak interactions. This asymmetric patterning of weak and strong links is crucial to food web stability [@Paine1992; @McCann1998; @Neutel2002]. In a recent paper on an aquatic food web it was further highlighted that temporal changes in ecosystem functioning could only be predicted using a weighted food web structure [@Kortsch2021]. Hence, in order to assess the stability and functioning of a food web, it is important to first determine link weights.

Several methodologies have been applied to estimate IS in food webs, where the quantity and quality of the data mostly determine which approach is the most convenient [@Berlow2004]. Approaches include experimental methods combined with dynamic modelling [@Emmerson2004; @Carrara2015], measurements of species abundances through time [@Fahimipour2014; @Chang2021], and estimation of metabolic rates and biomass of all species in the community [@Neutel2014]. However, these types of methods require large experimental set-ups and parameterisations restricting the analyses to smaller networks (e.g., approximately 10 species or less). Other methods based on allometric scaling relationships and biomass information [@Kortsch2021; @Gauzens2019] can be applied to larger networks with less data requirements, but this comes at the expense of precision in the predictions. For even larger food webs composed of nearly 1000 species and more than 10000 interactions, only methods with even less data requirements are feasible. One of these methods, proposed by @Pawar2012, combines data on consumer and resource body masses and consumer search space (interaction dimensionality) to obtain IS estimates for each pairwise predator-prey interaction. An advantage of this method is that it can be applied without information on species biomass.

Diverse food webs (e.g., terrestrial, lake, marine) from various geographic locations have been studied worldwide, including marine food webs in polar regions [@Carscallen2012; @Santana2013; @Kortsch2019; @Pecuchet2022]. Studies from the Arctic show that marine food web properties (e.g., connectance and modularity) are constrained by sea ice and seawater temperature [@Kortsch2019; @Pecuchet2022]. Likewise for the Southern Ocean it was shown that sea ice has impacts on local (Ross Sea) food web structure [@Rossi2019]. For example, in the Ross Sea, sea-ice break up introduced a surge of sympagic food, which led to simpler food webs with reduced intraguild predation, potential disturbance propagation and increased vulnerability to biodiversity loss [@Rossi2019].

The Weddell Sea is expected to be one of the last regions of the Southern Ocean to experience the consequences of climate change due to its extensive sea ice cover and ocean currents [@Teschke2021] resulting in less sea surface warming compared to other areas of the Southern Ocean. Generaly, the Southern Ocean plays an important role in driving global thermohaline circulation and ventilating the global abyssal ocean because it generates a considerable part of the Antarctic Bottom Water [@Fahrbach2009]. Because of these environmental characteristics, the Weddell Sea may serve as a refuge for Antarctic species which depend on sea ice (e.g. krill, emperor penguin, Weddell seal) or have low heat tolerance (e.g., most notothenoid fishes) due to their adaptations to freezing temperatures [@Griffiths2017]. While essential large-scale hydrodynamic relationships are relatively well-known for this region [@deSteur2019], information on the current distribution, abundance and sensitivity to climate change is only partially known for a few species (e.g., emperor penguin) [@Houstin2022].

The Weddell Sea food web is highly complex, comprising 488 species and 16200 predator-prey interactions [@Jacob2011]. To better understand species roles related to food web stability, Jacob et al. (2011) performed secondary extinctions and found that the removal of small to medium-sized, and not large, organisms caused a cascade of secondary extinctions. These findings highlighted the relative importance of predators, rather than prey, for the architecture, functioning and stability of the Weddell Sea food web, which coincides with findings from recent meta-analyses in empirical complex food webs [@Brose2019; @Perkins2022]. @Brose2019 considered this food web in a meta-analysis context showing that high predator-prey body-mass ratios are found for predator groups with specific trait combinations, including small vertebrates and large swimming or flying predators. These trait combinations generate weak interactions that stabilize communities against perturbations maintaining ecosystem functioning.

In this study, we aim to go beyond a purely topological (presence/absence) assessment of who eats whom in the Weddell Sea ecosystem by providing a quantitative analysis of the trophic network structure. We aim to analyse the species' role for the structure and stability of the food web. To achieve this, we: 1) estimated the strength for each interaction, 2) characterised species' role considering both weighted and unweighted properties, and 3) analysed species’ roles related to the stability of the food web.

@@

The high Antarctic Weddell Sea shelf is situated between 74 and 78ºS, stretching approximately 450 km from East to West (Figure 1). Water depth varies between 200 and 500 meters, and shallower areas are covered by continental ice, which forms the coastline along the eastern and southern parts of the Weddell Sea. The shelf area contains a complex three-dimensional benthic habitat with large benthic biomasses, intermediate to high diversity in comparison to benthic boreal communities and a spatially patchy distribution of organisms [@Dayton1990; @Teixido2002].

@@

The Weddell Sea food web was retrieved from the GlobAL daTabasE of traits and food Web Architecture (GATEWAy, version 1.0) of the German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig [@Brose2018]. In addition to predator-prey interactions, the database contains information on other biological data such as mean body mass and movement type for each species in the food web. Furthermore, it incorporates information about the interaction itself, such as the dimension of the predator search space (2 or 3 dimensions). In its current form, the Weddell Sea food web comprises 490 species and 16041 predator-prey interactions and thus constitutes one of the most resolved food webs constructed to date [@Jacob2011].

Complex empirical food webs are compilations of species and their potential feeding interactions within a given area and time period. The species occurrence data to construct the Weddell Sea food web (Antarctica) was sampled between 1983 and 2005, and hence the Weddell Sea food web is representative for this time period. The diet composition for each species was observed between 2001 and 2004 through a combination of field observations and stomach content analyses (further details can be found in @Jacob2005). Some species, such as benthic grazers and suspension feeders had poor taxonomic resolution of their prey, therefore laboratory information about their size, behavior, and stable isotope signatures was used to infer their feeding habits [@Jacob2011]. As the data used to construct the Weddell Sea food web is sampled during the summer season due to the inaccessibility of ecological field sampling in Antarctica during the winter, the Weddell Sea food web is representative of the summer season.
Moreover, the species in the food web are representative of adult specimens, and were not categorized into ontogenetic life stages (i.e., larvae, juveniles, and adults). As a result, ontogenetic diet shifts cannot be addressed with this food web. The average body masses of species were either directly measured [@Jacob2005] or taken from published accounts for marine mammals and seabirds [@Brose2006a].
Due to the static nature of the Weddell Sea food web and its low temporal resolution, it is not possible to assess phenological mismatches nor species temporal turnover. Because of these methodological limitations, the nature of the Weddell Sea food web data may likely lead to overestimations of some trophic interactions and network properties (connectance), compared to realised networks at any given point in time and space.
The spatial extent of the food web is described in the Study area section (Figure 1).

@@

To estimate the strength of each pairwise interaction in the food web, we followed an approach proposed by @Pawar2012. The minimum data requirements are body mass of the consumer (predator) and resource (prey), and the interaction dimensionality (ID) classified as 2 or 3 dimensions. The ID is defined as the search space dimensions of the predator, which is also equivalent to the movement space of the prey. Thus, the ID is classified as 2D when both predator and prey move in 2D (e.g., both are benthic) or if a predator moves in 3D and a prey in 2D (e.g., pelagic predator on benthic prey). The ID is classified as 3D when both predator and prey move in 3D (e.g., both pelagic) or if the predator moves in 2D and the prey in 3D (e.g., benthic predator, pelagic prey) [@Pawar2012].

GATEWAy v.1.0 provides information on the mean body mass for consumers and resources, except for ‘detritus’ and ‘sediment’, and the dimensionality for the majority of the interactions, though the latter is missing in some cases (924 interactions). To complete the missing data on species ‘dimensionality’, we used information about the movement type of predators and prey included in GATEWAy.

@@

We obtained estimates for resource density and the search rate from the scaling relationships with the resource and the consumer mass, respectively [@Pawar2012]. The coefficients of such relationships, determined by ordinary least squares regression, vary with the ID. On one hand, resource density scales with resource mass as power-law with different exponents in 2D and in 3D. Since mean mass for resources ‘phytodetritus’ and ‘sediment’ were not available in GATEWAy, we considered the body mass of the smallest phytoplankton species (‘Fragilariopsis cylindrus’) as a proxy. This is justified by the fact that ‘phytodetritus’ and ‘sediment’ are mainly composed of dead or senescent phytoplankton reaching the seabed [@Wolanski2011]. On the other hand, search rate scales with consumer mass as power-law with exponents in 2D and in 3D [@Pawar2012]. The equations for estimating the search rate and the resource density are specified in the Supplementary Material.

To quantify the variability in IS estimates, we considered the uncertainties inherent in estimating the exponents via standard linear regression. In this regard, the exponents reported in @Pawar2012 can be viewed as the mean of a normal random variable, with the standard deviation given by the associated standard errors. We generated 1000 random samples and calculated an IS value for each sample, with the final outcome of a distribution of IS estimates for each interaction. Given that the IS distributions are right-skewed and bounded below zero, we characterized the central tendency using the median IS. To represent the distribution variability, we reported the interquartile range (Supplementary Material, Figure S1). This approach accounts for the propagation of uncertainty from the exponent estimates into the derived IS values, allowing us to systematically assess the robustness of conclusions based on these interaction metrics.

We fitted the IS distribution (i.e. medians for each interaction) to six candidate models (Exponential, Gamma, log-Normal, Normal, Power-law and Uniform), using maximum likelihood [@McCallum2008], and selected the best fitting model by computing the Akaike Information Criterion AIC [@Burnham2002].

@@

To characterise the role of each species in the food web, we considered unweighted and weighted food web properties (Figure 2). Unweighted properties are related to properties commonly used in qualitative food web studies and only describe the presence or absence of interactions without any information on link strength [@Martinez1991; @Dunne2002; @Borrelli2014]. In contrast, weighted properties capture the importance of a trophic interaction by considering its strength.

To assess species roles as a function of the weighted food web, we focused on mean IS defined as the average strength of all interactions for a given species. Further, we calculated three unweighted species properties: a) species degree, i.e., the sum of in- and out-going interactions ; b) trophic level ; and c) trophic similarity, i.e., the trophic overlap based on shared and unique resources and consumers. These metrics were chosen to assess species’ roles based on the unweighted network. The species degree has often been linked to a species importance for the structure and functioning within a food web, i.e. perturbations to high-degree species may therefore have more significant effects on the food web robustness than low-degree species [@Dunne2002a; references in @Cirtwill2018a]. The trophic level offers information about a species vertical position in the food web, and top predators at higher trophic levels and primary producers at the base are particularly important for food web dynamics through top-down and bottom-up control [references in @Cirtwill2018a]. Trophic similarity is an index of trophic overlap considering the set of prey and predators for a pair of species; it measures one of the most important aspects of a species’ niches, the trophic niche [@Martinez1991; @Williams2000].

Furthermore, we took species habitat affiliation into account, which describes the physical position of a species within the ecosystem. Species were categorised as: 1) benthic, if it lives on the seafloor; 2) pelagic, if it lives close to the surface; 3) benthopelagic, if it moves between and connects the aforementioned environments; 4) demersal, if it lives and feeds on, or near, the bottom of the sea; and 5) land-based, if the consumer is not strictly aquatic but feeds predominantly on marine species. Species habitat affiliations were retrieved from @Jacob2011.

To study the relationship between species mean IS (weighted property) and the unweighted species properties, we performed linear regression analyses between the log mean IS and each of the aforementioned unweighted properties. Thus, we considered the IS as the dependent variable and the given unweighted property as the independent variable, and obtained the coefficients (slope and intercept) for the linear model. Models were fitted using the least squares approach. We also explored the mean IS distribution with the species' habitat.

@@

To analyse the impact of species on food web stability, we performed extinction simulations deleting one species at a time, that is for every extinction, network size was reduced by one species only. After each extinction, we calculated the stability of the network minus the removed species (489 nodes) and compared it with that of the whole network (490 nodes in total).

To estimate stability, we utilized the mean of the real part of the maximum eigenvalue of the Jacobian (or community) matrix, employing random values for the elements of the Jacobian while preserving the predator-prey sign structure [@Saravia2022; @Allesina2008;  @Pawar2009]. The Jacobian matrix, denoted as $C$, is fundamental to understanding the population dynamics and inter-species interactions within an ecological network. Each element $c_{ij}$ of $C$ represents the effect of a change in the $j^{th}$ species’ density on the $i^{th}$ species, at equilibrium. When $i$ equals $j$, $c_{ij}$ reflects the dependence of the $i^{th}$ species on its own density, offering insights into intra-species interactions and self-regulation. The maximum eigenvalue describes the rate at which a small disturbance decays or amplifies over time in the vicinity of an equilibrium. Precisely, the real parts of the eigenvalues indicate the rate of exponential growth (if positive) or decay (if negative) of perturbations. Thus, a more negative index is indicative of a more stable food web with reduced probability of perturbation amplification; in other words the network is more resilient to disturbances.

We conducted 1000 simulations for the removal of each species, calculating the maximum eigenvalue for the network in each case. For each simulation, we compared this value against the median maximum eigenvalue obtained from 1000 simulations of the complete network, thus generating a distribution of differences. A positive difference indicates that the food web's stability is greater without the targeted species, suggesting that the species in question contributes to the network's instability. Conversely, a negative difference implies that the network is less stable without the species, indicating a stabilizing effect. Due to the variability in the estimation of the eigenvalues, we decided to consider that a substantial impact on stability was reached when the proportion of either negative or positive differences within this distribution exceeded 0.55, meaning that 55% of the time the difference in stability was positive or negative. A detailed description of the stability calculations can be found in the Supplementary Material (Figure S2).

To identify the species with the highest effect on food web stability and their characteristics, we plotted the results of each species' extinction and its effect on food web stability (i.e., the stability difference between the whole network and the network minus one species) against weighted (interaction strength), and the three unweighted properties (trophic level, degree, and trophic similarity), and species habitat affiliation.

@@

The statistical distribution that best fitted the empirical interaction strength distribution of the Weddell Sea food web was a ‘log-Normal’ due to the high proportion of weak interactions and the existence of a few strong interactions (Figure 3, Table S3).

Here it is important to note that we observed an important variability when estimating each IS. The interactions that showed the greatest variability, represented by a wide interquartile range, were those where consumer and resource body masses were very small (approximately 1e-15 kg). The frequency distribution of the interquartile range is presented in the Supplementary Material (Figure S1).

@@

We found that the species’ mean IS (weighted property) shows different relationships with the unweighted properties analysed (Figure 4A-D). In this regard, there is a positive relationship between IS and trophic level, i.e., the higher the trophic level of the species, the higher its mean IS. We also found a significant but less evident positive relationship with species degree. Contrary, there was no significant relationship between mean IS and trophic similarity. Considering species habitat affiliation, the “Benthopelagic” and “Pelagic” categories contained the two species with the highest mean IS, the Orca Orcinus orca and the Colossal Squid Mesonychoteuthis hamiltoni, respectively. However, the majority of the species with relatively higher IS belonged to the “Demersal” and “Land-based” habitats groups. Species inhabiting the benthic realm showed the lowest mean IS (Figure 4D).

@@

Our extinction analyses showed that the majority of species had no significant impact on food web stability after being removed (Figure 5). Most of the species (yellow points in Figure 5) did not change the stability of the network considerably after being removed, except for a few species (purple points in Figure 5). Only 12 out of 490 species (2.45%) gave rise to substantial changes in the food web’s stability after their removal (Table 1). Network stability increased after the removal of most of these species, i.e., these species have a negative effect on stability. Only two species significantly decreased network stability after being removed, the minke whale Balaenoptera acutorostrata and the sea urchin Aporocidaris milleri. These two species have a positive effect on network stability.

After exploring the stability difference against the species properties (Figure 5), we found that the species that generated a substantial impact on the stability of the food web were characterised by: 1) high mean IS; 2) mid to high trophic levels (TL > 3.3); 3) relatively high number of interactions (Degree > 25); and 4) mid to low trophic similarity (TS < 0.13). Habitat wise, species with a significant impact on stability were present in all habitats. Table 1 shows the results for the species with highest impact on the food web stability.

@@

Our analyses show that the distribution of species IS at the network level is asymmetric, i.e., the Weddell Sea food web contains many weak interactions and only a few strong ones. This finding is consistent with many previous theoretical and empirical studies [e.g. @McCann1998; @Neutel2002; @Emmerson2004; @Wootton2005; @Kortsch2021]. The asymmetric distribution of IS in food webs has been interpreted as an explanation for the persistence of complex communities in nature [@Bascompte2005; @Allesina2015; @Nilsson2016]. Here we show that this pattern is also prevalent in the Weddell Sea, one of the most complex food webs to date, comprising 490 species and 16041 predator-prey interactions. This finding validates the method we used.

@@

```{r, out.width = "12cm", echo = FALSE, fig.cap = "Frequency distribution of (median) interaction strengths for the Weddell Sea food web. The distribution was best fitted to a 'log-Normal' model. Total number of interactions = 16041."}
knitr::include_graphics("Fig3_IntDist_sim.png")
```
